---
title: "TidyVerse Create and Extend (Vignette)"
author: "Joyce Aldrich"
date: "2023-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction


In this assignment, I will use some handy functions in `dplyr` package to tidy data.

 * dplyr: a set of tools for efficiently manipulating datasets
 
```{r 0, message=FALSE, warning=FALSE}
library(dplyr)
library(readr) # read csv file

```


## Dataset

The dataset in this project is called 'Student Performance' from <https://www.kaggle.com/datasets>  
The description of data is as below:

    "gender" - the gender of the student
    "race/ethnicity" - the student's racial or ethnic background
    "parental level of education" - the highest level of education attained by the student's parent(s) or guardian(s)
    "lunch" - whether the student receives free or reduced-price lunch (yes/no)
    "text preparation course" - whether the student completed a test preparation course (yes/no)
    "math score" - the student's score on a standardized mathematics test
    "reading score" - the student's score on a standardized reading test
    "writing score" - the student's score on a standardized writing test

## readr::read_csv 

I use `read_csv` function from `readr` to import csv file.
```{r}
student_exam <- read_csv("/Users/joycealdrich/Documents/SPS Data Science/Data 607/Vignette/student performance.csv")
head(student_exam)
```

## dplyr:rename

the `rename`function to change the column from old to a new name. I do the following:

```{r}
student_exam_rename <- rename( student_exam,
        race_ethnicity = `race/ethnicity`,
        parental_level_of_education =  `parental level of education` ,
        test_preparation_course = `test preparation course`,
        math_score = `math score`,
        reading_score = `reading score`,
        writing_score = `writing score`)
```
    

## dplyr:mutate

`mutate` function adds new variables into the dataframe and presevers the existing ones. For example, if I want to calculate the score of three exams 'math', 'reading' and 'writing' and the average score of these three exams and retain 'math', 'reading' and 'writing', I do the following:

```{r}

student_exam_mutate <- mutate( student_exam_rename, total_score = math_score + reading_score +writing_score, average= round((math_score + reading_score +writing_score)/3, digits = 2))

student_exam_mutate

```

## dplyr:select
the `select` function is straightforward, I can select the columns that I want. If I want to drop columns such as the 'lunch' and  ‘parental_level_of_education’, the select function is also handy.

```{r}
student_exam_select <- select(student_exam_mutate, -lunch, -parental_level_of_education)

student_exam_select
```



## dplyr:filter

the `filter` function is straightforward, I can exract rows that meet logical criteria. If I wound like to find the students whose math score and writing score are above 80 then the `filter` function is also handy.
```{r}
student_exam_filter <-  filter (student_exam_mutate, math_score >= 80 & writing_score >= 80)

student_exam_filter
```


